$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-05 paper/paper.tex
Shorten substitution proof, move the long version to the appendix

@@

\autoref{thm:subst-generalization} proves a generalized version of substitution, where the substitution $\subst{P}{j}{i}$ is on any variable $i$.
\autoref{thm:substitution} instantiates the generalized version to the concrete case where $i$ is $\constr{0}$, as required by subject reduction.
\todo{Structure the theorem (in general the long ones) a bit more. Use itemize or similar. It's a blob of text.}
\begin{nitheorem}
  \label{thm:subst-generalization}
  Let process $P$ be well-typed such that $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$,
  Let $\ni_i m$ be a \textsc{VarRef} $\contains{\gamma}{\Gamma_i}{i}{t}{m}{\Gamma}$ for some $m$ and $\Gamma$.
  Let $\ni_i n$ be a \textsc{VarRef} $\contains{\gamma}{\Psi_i}{i}{t}{n}{\Psi}$ for some $m$ and $\Psi$.
  Let $\Gamma$ and $\Psi$ be related such that $\opctx{\Gamma}{\Delta}{\Psi}$ for some $\Delta$.
  Let $\Delta$ have a usage annotation $\lz$ at position $i$.
  Let $\ni_j m$ be a \textsc{VarRef} $\contains{\gamma}{\Gamma_j}{j}{t}{m}{\Gamma}$ for some $\Gamma_j$.
  Let $\ni_j n$ be a \textsc{VarRef} $\contains{\gamma}{\Psi_j}{j}{t}{n}{\Psi}$ for some $\Psi_j$.
  Then substituting $i$ with $j$ in $P$ will be well-typed such that $\types{\gamma}{\Gamma_j}{\subst{P}{j}{i}}{\Psi_j}$.
  Refer to \autoref{fig:subst} for a diagramatic representation.
\end{nitheorem}

@@

\begin{nidefinition}
  For convenience, we use $\containsusage{\Gamma}{i}{x}{\Delta}$ to stand for $\contains{\gamma}{\Gamma}{i}{t}{x}{\Delta}$ for some $\gamma$ and $t$.
\end{nidefinition}

\todo{I don't think a reviewer from CONCUR most likely will nto get the proof. Appendix! Replace it with a proof sketch where you comment on the diagram as well.}
\begin{proof}
  By induction on the derivation $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  \begin{itemize}
    \item
      For constructor $\constr{end}$ we get $\Gamma_i \equiv \Psi_i$.
      From $\Delta_i \equiv \lz$ follows that $m \equiv n$.
      Therefore $\Gamma_j \equiv \Psi_j$ and $\constr{end}$ can be applied.

    \item
      For constructor $\constr{chan}$ we proceed inductively, wrapping arrows $\ni_i m$, $\ni_j m$, $\ni_i n$ and $\ni_j n$ with $\suc$.

    \item
      For constructor $\constr{recv}$ we must split $\Delta$ to proceed inductively on the continuation.
      Observe that given the arrow from $\Gamma_i$ to $\Psi_i$ and given that $\Delta$ is $\lz$ at index $i$, there must exist some $\delta$ such that $\opsquared{m}{\delta}{n}$.
 l     \begin{itemize}
        \item
          If the input is on the variable being substituted, we split $m$ such that $\opsquared{m}{\li}{l}$ for some $l$, and construct an arrow $\containsusage{\Xi_i}{i}{l}{\Gamma}$ for the inductive call.
          Similarly, we construct for some $\Xi_j$ the arrows $\containsusage{\Gamma_j}{j}{\li}{\Xi_j}$ as the new input channel, and $\containsusage{\Xi_j}{j}{l}{\Gamma}$ for the inductive call.
        \item
          If the input is on a variable $x$ other than the one being substituted, we construct the arrows $\containsusage{\Xi_i}{i}{m}{\Theta}$ (for the inductive call) and $\containsusage{\Gamma}{x}{\li}{\Theta}$ for some $\Theta$.
          We then construct for some $\Xi_j$ the arrows $\containsusage{\Gamma_j}{x}{\li}{\Xi_j}$ (the new output channel) and $\containsusage{Xi_j}{j}{m}{\Theta}$ (for the inductive call).
          Given there exists a composition of arrows from $\Xi_i$ to $\Psi$, we conclude that $\Theta$ splits $\Delta$ such that $\opctx{\Gamma}{\Delta_1}{\Theta}$ and $\opctx{\Theta}{\Delta_2}{\Psi}$.
          As $\lz$ is a minimal element, then $\Delta_1$ must be $\lz$ at index $i$, and so must $\Delta_2$.
      \end{itemize}

    \item
      $\constr{send}$ applies the ideas outlined for the $\constr{recv}$ constructor to both the \textsc{VarRef} doing the output, and the \textsc{VarRef} for the sent data.

    \item
      For $\constr{comp}$ we first find a $\delta$, $\Theta$, $\Delta_1$ and $\Delta_2$ such that $\containsusage{\Xi_i}{i}{\delta}{\Theta}$ and $\opctx{\Gamma}{\Delta_1}{\Theta}$ and $\opctx{\Theta}{\Delta_2}{\Psi}$.
      Given $\Delta$ is $\lz$ at index $i$, we conclude that $\Delta_1$ and $\Delta_2$ are too.
      Observe that $\opsquared{m}{\delta}{\psi}$, where $\psi$ is the usage annotation at index $i$ consumed by the subprocess $P$.
      We construct an arrow $\containsusage{\Xi_j}{j}{\delta}{\Theta}$, for some $\Xi_j$.
      We can now make two inductive calls (on the derivation of $P$ and $Q$) and compose their results.
  \end{itemize}  
\end{proof}

>>>>

% Two column itemize
\usepackage{multicol}

@@

\autoref{thm:subst-generalization} proves a generalized version of substitution, where the substitution $\subst{P}{j}{i}$ is on any variable $i$.
\autoref{thm:substitution} instantiates the generalized version to the concrete case where $i$ is $\constr{0}$, as required by subject reduction.

@@

\begin{nitheorem}
  \label{thm:subst-generalization}
  Let process $P$ be well-typed such that $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  The substituted variable $i$ has a usage $m$ in $\Gamma_i$, and a usage $n$ in $\Psi_i$.
  Substitution will take these usages $m$ and $n$ away from $i$ and transfer them to the variable $j$ we are substituting for.
  In other words, there must exist some $\Gamma$, $\Psi$, $\Gamma_j$ and $\Psi_j$ such that:
  \begin{multicols}{2}
  \begin{itemize}
    \item $\contains{\gamma}{\Gamma_i}{i}{t}{m}{\Gamma}$
    \item $\contains{\gamma}{\Gamma_j}{j}{t}{m}{\Gamma}$
    \item $\contains{\gamma}{\Psi_i}{i}{t}{n}{\Psi}$
    \item $\contains{\gamma}{\Psi_j}{j}{t}{n}{\Psi}$
  \end{itemize}
  \end{multicols}
  Let $\Gamma$ and $\Psi$ be related such that $\opctx{\Gamma}{\Delta}{\Psi}$ for some $\Delta$.
  Let $\Delta$ have a usage annotation $\lz$ at position $i$.
  Then substituting $i$ with $j$ in $P$ will be well-typed such that $\types{\gamma}{\Gamma_j}{\subst{P}{j}{i}}{\Psi_j}$.
  Refer to \autoref{fig:subst} for a diagramatic representation.
\end{nitheorem}

\begin{proof}
  By induction on $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  \begin{itemize}
    \item
      For constructor $\constr{end}$ observe that $\Gamma_i \equiv \Psi_i$ and thus $\Gamma_j \equiv \Psi_j$.
      The details are found in \autoref{app:subst-generalization}.

    \item
      For constructor $\constr{chan}$ we proceed inductively.

    \item
      For constructors $\constr{recv}$, $\constr{send}$ and $\constr{comp}$ we must find $\Theta$ in \autoref{fig:subst} and split the diagram along its vertical axis to proceed by induction.
      The details are found in \autoref{app:subst-generalization}.
  \end{itemize}  
\end{proof}
